Investigating magnetic dynamo evolution with \TESS\ field dwarfs

### 2017-10-06 ms.tex
adding Derek's corrections

\documentclass[letterpaper,12pt,preprint]{hack_aastex}

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky.
% Figure out how to incorporate the target list.

\input{dfm_stylez}
\pagestyle{myheadings}
\markright{\textsf{\footnotesize %
    Investigating rotational evolution with TESS FFI field dwarfs}}

@@

\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\tess}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\RAVE}{{\it RAVE}}
\newcommand{\rave}{{\it RAVE}}
\newcommand{\hermes}{{\it HERMES}}
\newcommand{\HERMES}{{\it HERMES}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

\newcommand{\racomment}[1]{{\color{red}#1}}

\newcommand{\columbia}{1}
\newcommand{\ww}{2}
\newcommand{\cca}{3}
\newcommand{\florida}{4}
\newcommand{\princeton}{5}
\newcommand{\nsf}{4}
\newcommand{\simons}{2}
\newcommand{\hubble}{7}

@@

\title{Investigating magnetic dynamo evolution with TESS field dwarfs}

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars.
To answer these questions {\bf we propose to provide a catalog of rotation
periods for \TESS\ Full Frame Image (FFI) stars.}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the Kepler rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measureable rotation periods
to less than around 15 days.
The longer time coverage in the CVZz will allow us to study the rotation
period bimodality in K and M dwarfs, since the bimodality appears at longer
rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

\ssfigure{Davenport.png}{0.5}{%
From Davenport (2017).
    Rotation period as a function of effective
temperature for the full McQuillan et al., (2014) Kepler sample in black, and
the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
blue. Contaminating giants have been removed from the blue sample and the
rotation period bimodality is revealed to exist across all temperatures shown.
    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).
\label{fig:davenport}
}

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^1/2$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

We have identified 34 stars in 17 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS
% \footnote{Based on {\tt tvguide}.}
and are already in the Candidate Target List.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have similar ages and measuring their rotation periods will
therefore provide a test of gyrochronology.
{\bf We will use the rotation periods of comoving stars identified in the
first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.}

\paragraph{Technical Feasibility}
The extraction of a rotation period from a light curve can be as simple as
computing a Lomb-Scargle periodogram or, in cases where the signal is less
clear, can be inferred via modeling the correlation between data points.
This latter approach could involve either computing an autocorrelation
function \citep[\eg][]{mcquillan2013}, or fitting a Gaussian process to the
time series \citep{angus2017, foreman-mackey2017}.
In cases where signals have long periods and low amplitudes, care is needed to
separate real astrophysical signal from instrumental systematics.
The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the TESS FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian process model
for the astrophysical variation.
Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{foreman-mackey2017} to scale the computations to the size of
TESS FFI datasets.
The precision of existing photometric de-trending methods degrades in crowded
fields \citep[for example,][]{luger2017}.
However, to make robust measurements of rotation periods, we do not need
absolutely calibrated light curves.
Therefore, in crowded fields, we will apply a difference imaging method that
was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
variations of crowded sources.
Unlike standard difference imaging methods, this procedure does not require a
reference image.
Instead, a causal data-driven model is built to predict the time series in
every pixel taking systematic effects into account and the residuals between
the observations and the model predictions provide an estimate of the
astrophysical variability in each pixel.
We will tune this method to preserve rotation signals and apply it to detect
rotation across the FFIs.

@@

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for around 30,000 high priority FFI targets and a catalog
of rotation periods for lower priority FFI stars using a combination of
Lomb-Scargle periodogram and autocorrelation techniques.
Finally, we will release a catalog of rotation periods for all two-minute
cadence targets.
The full catalog will likely contain several million rotation periods.
We will release code for inferring stellar ages from rotation periods, based
on a new gyrochronology relation calibrated using \TESS\HERMES spectra, \Gaia\
kinematics and comoving pairs.
We will also release an open source python package for measuring rotation
periods of both FFI and two-minute cadence targets.

>>>>

\documentclass[letterpaper,12pt,preprint]{hack_aastex}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\tess}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\RAVE}{{\it RAVE}}
\newcommand{\rave}{{\it RAVE}}
\newcommand{\hermes}{{\it HERMES}}
\newcommand{\HERMES}{{\it HERMES}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\racomment}[1]{{\color{red}#1}}

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky. ~ 15 million, call it 10.
% Figure out how to incorporate the target list.
% Get detection probabilities for stars on the list
% Apply Derek's corrections
% Calculate the distance at which you get a ~ 15th mag at ~1kpc
% cite the sullivan paper.
% Mention vetting.

\input{dfm_stylez}
\pagestyle{myheadings}
\markright{\textsf{\footnotesize %
    Investigating rotational evolution with \TESS\ FFI field dwarfs}}

@@

@@

\title{Investigating magnetic dynamo evolution with \TESS\ field dwarfs}

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars
\citep{vansaders2016}.
To answer these questions {\bf we propose to provide a catalog of rotation
periods for \TESS\ Full Frame Image (FFI) stars.}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The \TESS\ FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the \TESS\ FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
to less than around 15 days.
The longer time coverage in the CVZ will allow us to study the rotation
period bimodality in K and M dwarfs, since the bimodality appears at longer
rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

\ssfigure{Davenport.png}{0.5}{%
From Davenport (2017).
    Rotation period as a function of effective
temperature for the full McQuillan et al., (2014) \Kepler\ sample in black, and
the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
blue. Contaminating giants have been removed from the blue sample and the
rotation period bimodality is revealed to exist across all temperatures shown.
    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).
\label{fig:davenport}
}

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

\citet{oh2016} identified over 10,000 comoving pairs of stars using proper
motions from the first \Gaia\ data release.
Of these, 34 dwarfs and subgiants in 17 pairs have \RAVE\ radial velocities
consistent with being comoving and are already in the Candidate Target List.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have similar ages and measuring their rotation periods will
therefore provide a test of gyrochronology.
An additional 5000 bright comoving pair candidates lie in the southern
ecliptic hemisphere and will be ideal FFI targets for gyrochronology studies.
{\bf We will use the rotation periods of comoving stars to quantify the
accuracy and precision of gyrochronology.}

\paragraph{Technical Feasibility}
The extraction of a rotation period from a light curve can be as simple as
computing a Lomb-Scargle periodogram or, in cases where the signal is less
clear, can be inferred via modeling the correlation between data points.
This latter approach could involve either computing an autocorrelation
function \citep[\eg][]{mcquillan2013}, or fitting a Gaussian process to the
time series \citep{angus2017, foreman-mackey2017}.
In cases where signals have long periods and low amplitudes, care is needed to
separate real astrophysical signal from instrumental systematics.
The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the \TESS\ FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian process model
for the astrophysical variation.
Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{foreman-mackey2017} to scale the computations to the size of
\TESS\ FFI datasets.
The precision of existing photometric de-trending methods degrades in crowded
fields \citep[for example,][]{luger2017}.
However, to make robust measurements of rotation periods, we do not need
absolutely calibrated light curves.
Therefore, in crowded fields, we will apply a difference imaging method that
was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
variations of crowded sources.
Unlike standard difference imaging methods, this procedure does not require a
reference image.
Instead, a causal data-driven model is built to predict the time series in
every pixel taking systematic effects into account and the residuals between
the observations and the model predictions provide an estimate of the
astrophysical variability in each pixel.
We will tune this method to preserve rotation signals and apply it to detect
rotation across the FFIs.

@@

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for around 30,000 high priority FFI targets rotation
periods point estimates for lower priority FFI stars using a combination of
Lomb-Scargle periodogram and autocorrelation techniques.
% Finally, we will release a catalog of rotation periods for all two-minute
% cadence targets.
The full catalog will contain several million rotation periods.
We will release code for inferring stellar ages from rotation periods, based
on a new gyrochronology relation calibrated using \TESS -- \HERMES spectra,
\Gaia\ kinematics and comoving pairs.
We will also release an open source python package for measuring rotation
periods of \tess\ FFI targets.
